---
title: "newvis"
author: "mayra"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Add Library

```{r}
library(dplyr)
library(ggplot2)
library(plotly)
library(reshape2)
```

# Import Dataset

```{r}
data_orii <- read.csv("heart.csv")
data_orii
```

# check duplicate row

```{r}
sum(duplicated(data_orii))
```

# Make a copy of the dataset

```{r}
data_cleaner <- data_orii
```

# checking Missing Value

```{r}
colSums(is.na(data_cleaner))
```

# show the structure of the data

```{r}
str(data_cleaner)
```

# change category column to factors

```{r}
#Change category column with char data type to factor data type

data_cleaner$Sex <- factor(data_orii$Sex, levels = c("M", "F"), labels = c("Male", "Female"))

data_cleaner$gender <- as.numeric(data_cleaner$Sex)

#changing data label based on the source
data_cleaner$ChestPainType <- factor(data_orii$ChestPainType, 
                             levels = c("ATA", "NAP", "ASY", "TA"), 
                             labels = c("Typical Angina", "Non-anginal Pain", "Asymptomatic", "Typical Atypical"))

#save encode column
data_cleaner$bloodSugar <-  data_orii$FastingBS

#change levels 0 to otherwise and 1 to blood sugar > 120 ml/dl (this information are based on the source of the data)
data_cleaner$FastingBS <- factor(data_orii$FastingBS, levels = c(0, 1), labels = c("Otherwise", "Blood sugar > 120ml/dl"))

#save heart disease column to outcome column
data_cleaner$outcome <- data_orii$HeartDisease

#change levels 1 to Had Heart Disease and 0 to No Heart Disease
data_cleaner$HeartDisease <- factor(data_orii$HeartDisease, levels = c(0, 1), labels = c("No Heart disease",  "Had Heart Disease"))

#change levels N tp No and Y to Yes
data_cleaner$ExerciseAngina <- factor(data_orii$ExerciseAngina, levels = c("N", "Y"), labels = c("No", "Yes"))

#change Resting ECG and ST_Slope to factor
data_cleaner$RestingECG <- as.factor(data_cleaner$RestingECG)
data_cleaner$ST_Slope <- as.factor(data_cleaner$ST_Slope)

#rechecking the dataset structure
str(data_cleaner)
```

# summary of the dataset

```{r}
summary(data_cleaner) #checking any zero values
```

# clean the dataset

```{r}
#Attributes that has unusual 0 values or under 0 values (Oldpeak, RestingBP, cholesterol)
data_oldpeak <- data_cleaner %>%
  filter(Oldpeak < 0) #to check the minus data in data old peak

data_BPnotNol <- data_cleaner %>%
  filter(RestingBP > 0)
data_cholestrol <- data_cleaner %>%
  filter(Cholesterol != 0)

#replace 0 or under 0 values with median value of the data

data_cleaner$RestingBP_T <- ifelse(data_cleaner$RestingBP == 0, 
                                 mean(data_BPnotNol$RestingBP, na.rm = TRUE), data_cleaner$RestingBP)
data_cleaner$Oldpeak_T <- ifelse(data_cleaner$Oldpeak < 0, 0, data_cleaner$Oldpeak)
data_cleaner$Cholesterol_T <- ifelse(data_cleaner$Cholesterol == 0, 
                                     mean(data_cholestrol$Cholesterol, na.rm = TRUE), 
                                     data_cleaner$Cholesterol)  

#Remove column that we don't need
data_cleaner$Oldpeak <- NULL
data_cleaner$RestingBP <- NULL
data_cleaner$Cholesterol <- NULL

data_cleaner
summary(data_cleaner)
colSums(is.na(data_cleaner))
```

# make new attribute

```{r}
#make new attribut for age, cholesterol, oldpeak, restingBP and maxHR
data_final <- data_cleaner %>% #copy dataset with new attribute (as the finalal dataset)
  mutate(
    AgeCategory = case_when(
      Age < 30 ~ "< 30",
      Age >= 30 & Age < 45 ~ "30-44",
      Age >= 45 & Age < 60 ~ "45-59",
      Age >= 60 & Age < 75 ~ "60-74",
      Age >= 75 ~ "75 or Older"
    ),
    Cholestrol_Cat1 = case_when(
      Cholesterol_T < 200 ~ "Normal",
      Cholesterol_T >= 200 & Cholesterol_T < 240 ~ "At Risk",
      Cholesterol_T >= 240 ~ "High"
    ),
    Oldpeak_Cat = case_when(
      Oldpeak_T >= 0 & Oldpeak_T < 1~ "Healthy Heart",
      Oldpeak_T >= 1 & Oldpeak_T < 2 ~ "Potential Heart Issue",
      Oldpeak_T >= 2 ~ "Strong Coronary Heart disease"
    ),
    RestingBP_Cat = case_when(
      RestingBP_T < 120 ~ "Normal",
      RestingBP_T >= 120 & RestingBP_T < 130 ~ "Pre-Hypertensive",
      RestingBP_T >= 130 ~ "Hypertension",
    ),
    MaxHR_P = 220 - Age,
    HR_percent = MaxHR / MaxHR_P * 100,
    MaxHR_Cat = case_when(
      HR_percent < 60 ~ "Low",
      HR_percent >= 60 & HR_percent < 100 ~ "Normal",
      HR_percent >= 100 ~ "High"
      )
  )


#change the new attribute data type to factor
data_final <- data_final %>%
  mutate(
    RestingBP_Cat = factor(RestingBP_Cat, 
                             levels = c("Normal", "Pre-Hypertensive", "Hypertension")),
    Cholestrol_Cat1 = factor(Cholestrol_Cat1,levels = c("Normal", "At Risk", "High")),
    MaxHR_Cat = factor(MaxHR_Cat, levels = c("Low", "Normal", "High"))
    )

#Recheck missing value
colSums(is.na(data_final))

```

# Barchart for worldwide Non Comunicable Disease Death Cause

```{r}
#Make a new data frame for non communicable disease death rate
ncD_death <- data.frame(
  disease = c("Heart Disease","Covid 19", "Diabetes", "Stroke", "Chronic Respiratory Disease"),
  deaths = c(20.5, 8.00, 6.7, 7, 3.5)
)

#Sort disease in decreasing order
ncD_death_sort <- ncD_death[order(ncD_death$deaths, decreasing = TRUE),]

ncD_death_sort

#make the plot
ggplot(ncD_death_sort, aes(y = reorder(disease, deaths), x = deaths, fill = disease)) +
              geom_col(position = "dodge", width = 0.6)+
              geom_text(aes(label = deaths), 
              hjust = -0.1) +
              labs(title = "Non-Communicable Disease Death Rate",
                   x = "Death Rate (in Million)",
                   y = "Disease") +
              scale_fill_manual(
                values = c(
                  "Heart Disease" = "#f44336",
                  "Stroke" = "#bdbdbd",
                  "Covid 19" = "#bdbdbd",
                  "Diabetes" = "#bdbdbd",
                  "Chronic Respiratory Disease" = "#bdbdbd"
                  )
              ) +
              theme(
                panel.background = element_rect(fill = "white"),
                legend.position = "none"
                ) +
  coord_cartesian(clip = "off") 
```

# Piechart : cholesterol Status Percentage

```{r}
#make new dataframe that count and filter cholesterol only
sum_chol<- data_final%>%
  count(Cholestrol_Cat1) %>%
  mutate(perc = n/sum(n) * 100)

#change category color
custom_c= c("Normal"="#FFF89A","Risk" = "#F6BD60","High" = "#FF4500")

#make the plot
plot_ly(sum_chol,labels = ~Cholestrol_Cat1, values = ~perc, type = 'pie',
        textinfo = 'label+percent',marker = list(colors = custom_c))%>%
  layout(
    title = "cholesterol status diagram",
    xaxis = list(title = "Cholesterol"),
    yaxis = list(title = "Persentage", ticksuffix = "%"),
    legend = list(title = list(text = "Cholesterol Type"))
  )

```

# Piechart : Percentage of Patient with Heart Disease Status based on High Level cholesterol Population

```{r}
# Create a summary of heart disease count and percentage among individuals with high cholesterol
summary_HD <- data_final %>% 
  filter(Cholestrol_Cat1 == "High")%>%
  count(HeartDisease) %>%
  mutate(perc = n/sum(n) * 100)

#make the plos based onn the data summary
plot_ly(summary_HD, labels = ~HeartDisease, values = ~n, type = 'pie',textinfo = 'label+percent',
            marker = list(colors = c("No Heart disease" = "#A8DADd", "Had Heart Disease" ="#f44336")))%>%
      layout(
        title = "Patient Heart Disease Percentage in High Cholesterol Population",
        xaxis = list(title = "Heart Disease"),
        yaxis = list(title = "Persentage", ticksuffix = "%"),
        legend = list(title = list(text = "Heart Disease"))
      )

```

# Correlation Map : Heart Disease Factor

```{r}
#create a dataframe that only constains number
datas_num<- data_final%>% filter(Cholestrol_Cat1 == "High")
datas_num <- datas_num[c("Age", "RestingBP_T", "Cholesterol_T","gender", "MaxHR", "Oldpeak_T", "outcome", "bloodSugar")]

datas_num

# Calculate the correlation for all numeric variables
corr_num <- cor(datas_num)

# Order the variables by their correlation strength with the 'outcome' variable (from highest to lowest)
ordered <- names(sort(corr_num["outcome",], decreasing = TRUE))

# Reorder the correlation based on the variables most correlated with 'outcome'
order_corr <- corr_num[ordered, ordered]

ggplot(data = melt(order_corr), aes(Var1, Var2, fill = value)) +
  geom_tile() +
  labs(x = "HeartDisease Factors", y = "Heartdisease Factors", fill = "Correlation") +
  scale_fill_gradient(low = "#ffcccc", high = "#f44336") +
   theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    panel.background = element_rect(fill = "white")
  )
```

# Barchart (for Shiny) : Amount Patient with Cholesterol

```{r}
#make a plot only for Cholesterol Distribution
custom_c= c("Normal"="#FFF89A","At Risk" = "#F6BD60","High" = "#FF4500")
ggplot(data_final, aes(x =Cholestrol_Cat1, fill = Cholestrol_Cat1)) +geom_bar(position = "dodge")+
  scale_fill_manual(values = custom_c) +
  labs(
    title = "Amount of Patient with Cholesterol",
    x= "cholesterol Category",
    y = "Amount of Patient",
    fill = "cholesterol Level"
  )
```

# Barchart (For shiny) : Amount of Patient with Heart Attack based on High Level cholesterol

```{r}
#create a summary based of heart disease and cholesterol count
hd_clo_data <- data_final %>%
  count(Cholestrol_Cat1, HeartDisease)%>%
  mutate(perc = n / sum(n) * 100,Combined = interaction(Cholestrol_Cat1, HeartDisease, sep = " - ")
  )

customCol <- c(
  "Normal - Had Heart Disease" = "#969696", "Normal - No Heart disease" = "#bdbdbd",
  "At Risk - Had Heart Disease" = "#969696", "At Risk - No Heart disease" = "#bdbdbd",
  "High - Had Heart Disease" = "#f44336", "High - No Heart disease" = "#ffcccc"
)
ggplot(hd_clo_data, aes(x =Cholestrol_Cat1, y =n, fill = Combined)) +geom_bar(stat = "identity", position = "dodge")+
  scale_fill_manual(values = customCol) +
  labs(
    title = "Amount of Patient with Heart Disease based on High Level cholesterol",
    x= "cholesterol Category",
    y = "Amount of Patient",
    fill = "cholesterol Level - Heart Disease"
  )
```

# Barchart : Percentage of Patient Blood Sugar that Has Heart Disease in High Level cholesterol Population

```{r}
#make a custom color for each bar in the plot
CustC <- c(
  "Otherwise - No Heart disease" = "#d9d9d9", "Blood sugar > 120ml/dl - No Heart disease" = "#ffcccc",
  "Otherwise - Had Heart Disease" = "#969696","Blood sugar > 120ml/dl - Had Heart Disease" = "#f44336"
)

#make a summary of heart disease and Fasting BS count among individual with High Cholesterol
summary_BSGB <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, FastingBS)%>%
  group_by(FastingBS)%>%
  mutate(perc = round(n / sum(n) * 100),Combined = interaction(FastingBS, HeartDisease, sep = " - ")
  )
summary_BSGB

#make a percentage plot based on the summary table
plot_ly(data = summary_BSGB, x = ~FastingBS,
        y = ~perc,
        color = ~Combined,
        colors=CustC, type = 'bar') %>%
  layout(
    title = "Blood Sugar Level divided by Heart disease",
    xaxis = list(title = "blood sugar level"),
    yaxis = list(title = "percentage", ticksuffix = "%"),
    barmode = 'stack',
    legend = list(title = list(text = "Heart Disease Status - Blood Sugar Level"))
  )
```

# Barchart : Percentage of Patient that Has Heart Disease based on Gender in High Level cholesterol Population

```{r}
# make a custom color for each bar in the plot
CustC <- c(
  "Male - No Heart disease" = "#ffcccc", "Male - Had Heart Disease" = "#f44336",
  "Female - No Heart disease" = "lightgray", "Female - Had Heart Disease"= "#969696" 
)

#make a summary based of heart disease and sex count among individuals that have high Cholesterol
sum_gender <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, Sex)%>%
  group_by(Sex)%>%
  mutate(perc = round(n / sum(n) * 100), Combined = interaction(Sex,HeartDisease, sep = " - ")
  )
sum_gender

#make the plot based on the summary table
plot_ly(data = sum_gender, x = ~Sex, y = ~perc, color = ~Combined, colors=CustC, type = 'bar') %>%
  layout(
    title = "Percentage Patient with Heart disease by Gender",
    xaxis = list(title = "Gender"),
    yaxis = list(title = "Percentage",ticksuffix = "%"),
    barmode = 'stack',
    legend = list(title = list(text = "Gender - Heart Disease Status"))
  )
```

# Barchart : Percentage of Patient that Has Heart Disease based on Blood Pressure in High cholesterol Population

```{r}
# make a custom color for each bar in the plot
cusColor <- c(
  "Normal - No Heart disease" = "#d9d9d9", "Pre-Hypertensive - No Heart disease" = "lightgray",
  "Hypertension - No Heart disease" = "#ffcccc",
  "Normal - Had Heart Disease" = "#969696","Pre-Hypertensive - Had Heart Disease" = "#969696",
  "Hypertension - Had Heart Disease" = "#f44336"
)

#make a summary based of heart disease and Resting BP count among individuals that have high Cholesterol
bp_HDBP_dataGB <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, RestingBP_Cat) %>%
  group_by(RestingBP_Cat) %>%
  mutate(perc = n / sum(n) * 100,Combined = interaction(RestingBP_Cat, HeartDisease, sep = " - ")
  )
bp_cholBP_dataGB

#make the plot based on the summary table
plot_ly(data = bp_HDBP_dataGB, x = ~RestingBP_Cat, y = ~perc, 
        color = ~Combined, colors=cusColor, type = 'bar') %>%
  layout(
    title = "Resting blood pressure levels divided by Heart disease status",
    xaxis = list(title = "Resting blood pressure levels"),
    yaxis = list(title = "percentage", ticksuffix = "%"),
    legend = list(title = list(text = "Resting blood pressure level - Heart disease status")),
    barmode = 'stack'
  )
```

# Barchart : Percentage of Patient with Heart Disease based on Age Category in High cholesterol Population.

```{r}
# make a custom color for each bar in the plot
custom_colo <- c("<30 Had heart Disease" = "#969696", "No Heart disease < 30"= "#d9d9d9",  
  "Had Heart Disease 30-44" = "#969696","No Heart disease 30-44"  = "lightgray",
  "Had Heart Disease 45-59" = "#f44336","No Heart disease 45-59"  = "#ffcccc",
  "Had Heart Disease 60-74" = "#f44336","No Heart disease 60-74"  = "#ffcccc",   
  "Had Heart Disease 75 or Older"= "#f44336", "No Heart disease 75 or Older"= "#ffcccc"
)

# Prepare the data
bp_agre_data <- data_final %>%
  filter(Cholestrol_Cat1 == "High") %>%
  count(AgeCategory, HeartDisease) %>%
  group_by(AgeCategory) %>%
  mutate(
    perc = n / sum(n) * 100,
    Combined = interaction(HeartDisease, AgeCategory, sep = " "))
bp_agre_data

#make the plot based on the summary table
plot_ly(data = bp_agre_data, x = ~AgeCategory, y = ~perc, color = ~Combined, 
        colors = custom_colo, type = 'bar'
) %>%
  layout(
    title = "Heart Disease Status based on Age in High Cholesterol Population",
    xaxis = list(title = "Umur"),
    yaxis = list(title = "percentage", ticksuffix = "%"),
    legend = list(title = list(text = "Age - Heart Disease Status")),
    barmode = 'stack'
  )
```

# Barchart : Percentage of Heart Disease based on Oldpeak Category in High cholesterol Population

```{r}
#make a summary based of heart disease and Oldpeak_Cat count among individuals that have high Cholesterol
sum_OP <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, Oldpeak_Cat)%>%
  group_by(Oldpeak_Cat) %>%
  mutate(perc = n / sum(n) * 100,Combined = interaction(Oldpeak_Cat, HeartDisease, sep = " - ")
  )
sum_OP

# make a custom color for each bar in the plot
custom_colorscl <- c("Healthy Heart - No Heart disease" = "#d9d9d9", 
                     "Potential Heart Issue - No Heart disease" = "lightgray", 
                     "Strong Coronary Heart disease - No Heart disease" = "#ffcccc",
                     "Healthy Heart - Had Heart Disease" = "#969696", 
                     "Potential Heart Issue - Had Heart Disease" = "#969696", 
                     "Strong Coronary Heart disease - Had Heart Disease" = "#f44336")

#make the plot based on the summary table
plot_ly(data = sum_OP, x = ~Oldpeak_Cat, y = ~perc, color = ~Combined, colors=custom_colorscl,
        type = 'bar') %>%
  layout(
    title = "oldpeak indicator divided by Heart disease status",
    xaxis = list(title = "oldpeak indicator"),
    yaxis = list(title = "percentage", ticksuffix = "%"),
    legend = list(title = list(text = "oldpeak indicator - Heart Disease status")),
    barmode = 'stack'
  )

```

# Barchart : Percentage of Heart Disease based on ST_Slope in High cholesterol Population

```{r}
# make a custom color for each bar in the plot
customCol<-c("No Heart disease - Down"="lightgray", "No Heart disease - Flat"="#ffcccc", "No Heart disease - Up"="#d9d9d9",
             "Had Heart Disease - Down"="#969696", "Had Heart Disease - Flat"="#f44336", "Had Heart Disease - Up"= "#969696")

#make a summary based of heart disease and ST_Slope count among individuals that have high Cholesterol
bp_stslop_data <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease,ST_Slope) %>%
  group_by(ST_Slope) %>%
  mutate(perc = n / sum(n) * 100,Combined = interaction(HeartDisease, ST_Slope, sep = " - "))
bp_stslop_data

#make the plot based on the summary table
plot_ly(data = bp_stslop_data, x = ~ST_Slope, y = ~perc, color = ~Combined, colors=customCol, type = 'bar'
) %>%
  layout(title = "ST slope divided by heart disease status ",
         xaxis = list(title = "ST slope"),yaxis = list(title = "percentage", ticksuffix = "%"),
         legend = list(title = list(text = "ST slope - Heart disease status ")),
    barmode = 'stack'
  )
```

# Barchart : Percentage of Heart Disease based on Type of Chest Pain in High cholesterol Population

```{r}
# make a custom color for each bar in the plot
custom_co <- c("No Heart disease - Typical Angina"= "lightgray", 
"No Heart disease - Non-anginal Pain"= "lightgray", "No Heart disease - Asymptomatic"= "#ffcccc",  
"No Heart disease - Typical Atypical"= "lightgray",  "Had Heart Disease - Typical Angina"     = "#969696",
"Had Heart Disease - Non-anginal Pain"   = "#969696",  "Had Heart Disease - Asymptomatic"       = "#f44336",  
"Had Heart Disease - Typical Atypical"   = "#969696"  
)

bp_type_data <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, ChestPainType) %>%
  group_by(ChestPainType) %>%
  mutate(perc = n / sum(n) * 100, Combined = interaction(HeartDisease, ChestPainType, sep = " - "))
bp_type_data

#make the plot based on the summary table
plot_ly(data = bp_type_data, x = ~ChestPainType, y = ~perc,color = ~Combined, colors=custom_co, type = 'bar'
) %>%
  layout(title = "chest pain type divided by Heart disease status",
         xaxis = list(title = "chest pain type"),
         yaxis = list(title = "percentage", ticksuffix = "%"),
         legend = list(title = list(text = "Heart disease status - types of chest pain")),barmode = 'stack'
  )
```

# Barchart : Percentage of Patient with Heart Disease that Has Exercise Angina In High level cholesterol Population

```{r}
# make a custom color for each bar in the plot
custom_co <- c(
  "No - No Heart disease" = "#d9d9d9", "Yes - No Heart disease" = "#ffcccc",
  "No - Had Heart Disease" = "#969696","Yes - Had Heart Disease" = "#f44336"
)

#make a summary based of heart disease and exercise angina count among individuals that have high Cholesterol
sum_HD_Angina <- data_final %>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, ExerciseAngina) %>%
  group_by(ExerciseAngina) %>%
  mutate(perc = round(n / sum(n) * 100),Combined = interaction(ExerciseAngina, HeartDisease, sep = " - ")
  )
sum_HD_Angina

#make the plot based on the summary table
plot_ly(sum_HD_Angina, x = ~ExerciseAngina,y = ~perc, color = ~Combined, colors=custom_co, type = 'bar') %>%
   layout(
    title = "Excersie Angina based on Heart disease status",
    xaxis = list(title = "Heart disease status"),
    yaxis = list(title = "percentage", ticksuffix = "%"),
    legend = list(title = list(text = "Heart Disease Status - Excersise Angina")),
    barmode = 'stack'
  )

```

# Barchart : Percentage of Patient with Heart Disease based on Heart Rate Level in High Level cholesterol Population.

```{r}
#make a summary based of heart disease and MAxHR_Cat count among individuals that have high Cholesterol
sum_HR <- data_final%>%
  filter(Cholestrol_Cat1=="High")%>%
  count(HeartDisease, MaxHR_Cat)%>%
  group_by(MaxHR_Cat)%>%
  mutate(perc = round(n / sum(n) * 100),Combined = interaction( MaxHR_Cat,HeartDisease, sep = " - ")
  )
sum_HR

# make a custom color for each bar in the plot
CustC <- c(
  "High - No Heart disease" = "lightgray", "Normal - No Heart disease" = "lightgray", "Low - No Heart disease" = "#ffcccc",
  "High - Had Heart Disease" = "#969696", "Normal - Had Heart Disease" = "#969696", "Low - Had Heart Disease" = "#f44336"
)

#make the plot based on the summary table
plot_ly(data = sum_HR, x = ~MaxHR_Cat, y = ~perc, color = ~Combined, colors= CustC, type = 'bar') %>%
  layout(
    title = "Maximum heart rate divided by Heart disease",
    xaxis = list(title = "maximum heart rate"),
    yaxis = list(title = "percentage", ticksuffix="%"),
    barmode = 'stack',
    legend = list(title = list(text = "Max Heart Rate - Heart Disease Status"))
  )
```
